### 2016-11-30 real/draft.tex
[draft] Introductory paragraphs edited, still in progress

%% CHANGE THIS DATE AS YOU MODIFY THE FILE
\newcommand{\vdate}{160510}
\pagestyle{myheadings}
\markboth{Version \vdate} {Version \vdate}
\parindent=0mm
\def\pp{\hfill\par \vspace{\baselineskip}}
%% \def\pp{}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{bm,amssymb,amsfonts,amsmath}  % bold math symbols, AMStex
\usepackage{graphicx}    % standard graphics
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\lleq}[1]{\label{#1} }
% TO REMOVE THE LABEL LETTERS FROM THE EQUATION DISPLAY, COMMENT OUT THIS LINE:
\renewcommand{\lleq}[1]{\label{#1} {\scriptstyle {\rm (#1)}} \hspace*{2ex} }
% \renewcommand{\labelenumi}{\arabic{section}.\arabic{subsection}.\arabic{enumi}}
% \renewcommand{\thesubsubsection}{\arabic{subsubsection}}

\newcounter{mnumi}
\newenvironment{mnumerate}{
  \begin{list}{\arabic{mnumi}.}
    {\usecounter{mnumi} \setlength{\itemsep}{0pt}
      \setlength{\leftmargin}{3ex}
    }
  }
  {\end{list}}

\newenvironment{mtemize}{
  \begin{list}{$\bullet$}
    {\setlength{\itemsep}{0pt}
     \setlength{\leftmargin}{3ex}
    }
  }
  {\end{list}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Bayesian prediction}

We first sketch the generic notation and relations for Bayesian
probability theory.
%
A particular experiment or simulation generates \textit{data} $D$
which will often come in the form of a list of $n$ measurements,
$D = \bmy = (y_1,y_2,\ldots,y_n)$. In order to understand or explain
this data, the observer constructs one or more \textit{models} or
\textit{hypotheses} $\hypo$. Central to this construction is the
specification of a \textit{likelihood} $p(\bmy\cond \bmtheta,\hypo)$
where $\bmtheta = (\theta_1,\theta_2,\ldots,\theta_K)$ is the set of
unknown parameters specified by the model. If and when the individual
measurements $y_i$ are independent, the joint likelihood is the
product
\begin{align}
  p(\bmy\cond \bmtheta) &= \prod_{i=1}^n p(y_i\cond \bmtheta).
\end{align}
At this level of analysis, the goal
is to find which values for these parameters provide a satisfactory
description of the data. In the Bayesian context, these parameter
values are best described in terms of the \textit{posterior}, which
is related to the likelihood by
\begin{align}
  \lleq{bth}
  p(\bmtheta\cond \bmy)
  &= \frac{p(\bmy\cond\bmtheta)\;p(\bmtheta)}
    {p(\bmy)}
    %= \frac{p(\bmy\cond\bmtheta)\;p(\bmtheta)}
    %{\int d\bmtheta'\, p(\bmy\cond\bmtheta')\;p(\bmtheta')}
\end{align}
where $p(\bmtheta)$, the so-called \textit{prior}, is the probability
assigned to the values of the parameters before any data was taken
into account and the \textit{evidence}
\begin{align}
  p(\bmy) &= \int d\bmtheta'\, p(\bmy\cond\bmtheta')\;p(\bmtheta'),
\end{align}
is fully determined by the likelihood and prior.

The scope of this paper is limited to the relationship between
simulated data generated by radiative transfer simulation codes.
Likewise, all parameters introduced and calculated relate strictly to
model descriptions of the simulation.
%
While Bayesian concepts and methods can be applied to actual
experimental data and related physics parameters, we shall not do so
here.

\hanslong{
Note that if the prior is uniform, it is independent of $\theta$ and
will cancel,
\begin{align}
  \lleq{bthu}
  p(\theta\cond \bmy)
  &= \frac{p(\bmy\cond\theta)} {\int d\theta'\, p(\bmy\cond\theta')}.
\end{align}
}

\hanslong{Bayesian prediction?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{TARDIS}
\fred{I don't think TARDIS is a good title, too specific}

We consider a computer simulation which, based on tunable input
parameters $\bmtheta$, generates $\npack$ photon packets, each with a
particular frequency $\nu_i$ and luminosity $\ell_i$. Together, these
frequencies and luminosities form the simulation data
$\bmy = \{ (\nu_i,\ell_i): i = 1,\ldots,\npack\}$.  We divide up the
frequency spectrum into bins
$\Delta_b = [\nu_b {-} \tfrac{1}{2} \Delta\nu \,,\, \nu_b {+}
\tfrac{1}{2} \Delta\nu], b = 1,2,\ldots, B$ of width $\Delta\nu$
centered on frequencies $\nu_b$. Let $n_b$ be the number of packets
having a frequency falling into a particular bin $b$, and let $\bml_b$
be the set of luminosities of those $n_b$ packets. The total
luminosity in bin $b$ is the sum of luminosities of those packets
whose frequency falls into $b$,
\begin{align}
  \ell_b = \sum_{i=1}^\npack \bm{1}_{\Delta_b}(\nu_i)\,\ell_i
\end{align}
where the indicator function
$\bm{1}_{[a,b]}(x) = 1$ when $x$ is in the interval
$[a,b]$ and 0 otherwise.

\section{Notation} \label{sec:notation}

\fred{Merge this section with previous section}

We consider a computer simulation that, given input parameters
$\bmtheta$ and a random-number seed, produces a set of simulated data
$ S = \{ (\nu_i, \ell_i): i=1\dots\npack \} = (\bmnu, \bml)$
consisting of $\npack$ packets, each with a frequency and a
luminosity. We distinguish between a random variable $L$ and its
concrete realization $\ell$.  To learn about $\bmtheta$, we need to
compare to data from a telescope available as luminosities in
frequency bins, $D = \{ ( \nu_b, \ell_b): b=1 \dots B\}$. The natural
estimate from a single simulation is to add up the luminosities of all
packets in a bin to estimate $\Lumtot_b$.  With a different seed but
identical $\bmtheta$, both the number of packets in the bin and their
luminosities are usually different. Our goal is to capture what is
known about $\Lumtot$ based on a \emph{single} simulation without recourse
to expensive repetitions or asymptotic approximations; i.e., we seek
$P(\Lumtot_b \cond \nu_b, \npack, \bmnu, \bml )$. In an actual physics
analysis, one would proceed to plug in the observed value $\ell_b$ and
perform an analysis to learn about $\bmtheta$. But here we focus on
the specific problem of $P(\Lumtot_b \cond \nu_b, \npack, \bmnu, \bml )$ and
hence omit $\bmtheta$ throughout.

Various features, for example absorption lines, usually make it
difficult to explicitly model the spectrum---the (total) luminosity as
a function of frequency. In a binned analysis, we can avoid these
difficulties associated to modeling $n_b$. Our approach, however,
requires an explicit expression of how the distribution of the
luminosity of a single packet in our simulation changes with $\nu$. It
is this crucial piece of information that allows us to extend previous
methods \fred{Refs: Zech, maybe others, too} in dealing with the
extreme cases when the number of packets in a bin, $n_b$, is one or
even zero.

>>>>

%% CHANGE THIS DATE AS YOU MODIFY THE FILE
\newcommand{\vdate}{161130}
\pagestyle{myheadings}
\markboth{Version \vdate} {Version \vdate}

\parindent=3ex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{bm,amssymb,amsfonts,amsmath}  % bold math symbols, AMStex
\usepackage{graphicx}    % standard graphics
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\lleq}[1]{\label{#1} }
% TO REMOVE THE LABEL LETTERS FROM THE EQUATION DISPLAY, COMMENT OUT THIS LINE:
\renewcommand{\lleq}[1]{\label{#1} {\scriptstyle {\rm (#1)}} \hspace*{2ex} }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Bayesian prediction}
\label{sec:bayp}

We first sketch the generic notation and relations for Bayesian
probability theory.
%
A particular experiment or simulation generates
\textit{data}
which will often come in the form of a list of $n$ measurements,
$\bmy = (y_1,y_2,\ldots,y_n)$. In order to understand or explain this
data, the observer constructs one or more \textit{models} or
\textit{hypotheses} $\hypo$. Central to this construction is the
specification of a \textit{likelihood} $p(\bmy\cond \bmtheta,\hypo)$
where $\bmtheta = (\theta_1,\theta_2,\ldots,\theta_K)$ is the set of
unknown parameters specified by the model. The likelihood is the
probability assigned by the model $\hypo$ to the data $\bmy$ when the
parameters $\bmtheta$ are fixed and known.
% 
If and when the individual measurements $y_i$ are independent, the
joint likelihood is the product
\begin{align}
  \lleq{bta}
  p(\bmy\cond \bmtheta) &= \prod_{i=1}^n p(y_i\cond \bmtheta).
\end{align}
At this level of analysis, the goal is to find which values for these
parameters provide a satisfactory description of the data. In the
Bayesian context, these parameter values are best described in terms
of the probability of the parameters given the data,
$p(\bmtheta\cond \bmy)$. This so-called \textit{posterior} is related
to the likelihood by Bayes' Theorem,
\begin{align}
  \lleq{bth}
  p(\bmtheta\cond \bmy)
  &= \frac{p(\bmy\cond\bmtheta)\;p(\bmtheta)}
    {p(\bmy)}
    %= \frac{p(\bmy\cond\bmtheta)\;p(\bmtheta)}
    %{\int d\bmtheta'\, p(\bmy\cond\bmtheta')\;p(\bmtheta')}
\end{align}
where $p(\bmtheta)$, the so-called \textit{prior}, is the probability
assigned to the values of the parameters before any data was taken
into account and the \textit{evidence}
\begin{align}
  \lleq{bti}
  p(\bmy) &= \int d\bmtheta'\, p(\bmy\cond\bmtheta')\;p(\bmtheta'),
\end{align}
is fully determined by the likelihood and prior. 
% 
Prediction of any ``future'' data $Y$ is made not by considering the
likelihood $p(Y\cond\bmtheta_{\rm max})$ using single best-fit
parameter values $\bmtheta_{\rm max}$, but by integrating over all
possible values of the $\bmtheta$ in terms of the posterior
$p(\bmtheta\cond n,\bml)$,
\begin{align}
  \lleq{btj}
  p(Y\cond \bmy) &= \int d\bmtheta\,p(Y\cond\bmtheta)\,p(\bmtheta\cond \bmy).
\end{align}
The above relations are true for any type of data and can be applied
to actual experimental data and the pertinent physical parameters. The
scope of this paper, however, is limited to the relationship between
simulated data generated by radiative transfer simulation codes.
Likewise, all parameters considered below relate strictly to model
descriptions of the simulation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{From single-simulation data to many-simulation prediction}

% \subsection{Output data from radiative transfer simulations}
\subsection{Simulation data}

We consider a single computer simulation which generates a luminosity
spectrum from $\npack$ photon packets, each with a particular
frequency $\nu_i$ and luminosity $\lum_i$. Together, these frequencies
and luminosities form the raw simulation data
$\bmy = \{ (\nu_i,\lum_i): i = 1,\ldots,\npack\}$.  We divide up the
frequency spectrum into bins of width $\Delta\nu$ centered on
frequencies $\nu_b$ resulting in bin intervals
$\Delta_b = [\nu_b {-} \tfrac{1}{2} \Delta\nu \,,\, \nu_b {+}
\tfrac{1}{2} \Delta\nu], b = 1,2,\ldots, B$. Let $n_b$ be the number
of packets having a frequency falling into a particular bin $b$, and
let the vector $\bml_b = \{(\ell_{b,j}): j = 1,\ldots,n_b\}$ be the
set of luminosities of those packets. The data set is correspondingly
cast into the more compact histogram form
\begin{align}
  \lleq{sdb}
  \bmy = \{(n_b,\bml_b): b = 1,\ldots,B\}
\end{align}
with loss of information limited to replacing the exact packet
frequencies $\nu_i$ by the appropriate $\nu_b$. The total luminosity
in bin $b$ is the sum of luminosities of those packets whose frequency
falls into $\Delta_b$,
\begin{align}
  \lleq{sdc}
  \lumtot_b 
  % &= \sum_{i=1}^\npack \bm{1}_{\Delta_b}(\nu_i)\,\ell_i
  &= \sum_{j=1}^{n_b} \lum_{b,j}.
\end{align}
% where the indicator function $\bm{1}_{[a,b]}(x) = 1$ when $x$ is in
% the interval $[a,b]$ and 0 otherwise.

\subsection{Prediction of luminosity spectrum}

Within a given model $\hypo$, the data $\bmy$ is only one of many
possible realisations of the corresponding random variables; for
example, starting the simulation with a different seed would result in
different output data. We denote the general random variables by the
corresponding upper-case symbols,
\begin{align}
  \bmL_b &= \{(\Lum_{b,j}): j = 1,\ldots,N_b\}\\
  \bmY &= \{(N_b, \bmL_b): b = 1,\ldots,B\},\\
  \Lumtot_b &= \sum_{j=1}^{N_b} \Lum_{b,j},
\end{align}
with corresponding likelihood $p(\bmY\cond\bmtheta)$ and posterior
$p(\bmtheta\cond\bmY)$ for any $\bmY$. The lower-case $(n_b,\bml_b)$
are strictly reserved for the simulated data at hand.

The goal of this paper is to illustrate how the single-simulation data
$\bmy$ can be used to predict the distribution of total bin
luminosities $Q_b, b = 1,\ldots,B$ if many simulations had been
carried out.

Various features such as absorption lines make the luminosity spectrum
highly variable. The model we develop below takes into account this
variability in a two-step process. First, the number of packets $N_b$
is modelled by means of a separate Poisson distribution
$p(N_b\cond\lambda_b)$ for each bin, where the parameter $\lambda_b$
will capture most of the variability of the spectrum.  Second, the
single-packet luminosity distribution is modelled by means of a packet
likelihood for $\Lum_{b,j}$ with packet parameters, a packet prior and,
of course, packet parameter posterior. The packet parameters, which
depend on the frequency bin, will be calculated in two ways: first by
drawing them from a ``parametrisation of the parameters'' across all
bins as shown in \refsec{example} and secondly by assuming that
\refsec{asymptotic}

Our approach
is to predict each bin's luminosity $Q_b$ separately using the same
model but allowing its parameters to vary independently in each bin.
Connections between them

function of frequency. In a binned analysis, we can avoid these
difficulties associated to modeling $n_b$. Our approach, however,
requires an explicit expression of how the distribution of the
luminosity of a single packet in our simulation changes with $\nu$. It
is this crucial piece of information that allows us to extend previous
methods in dealing with the extreme cases when the number of packets
in a bin, $n_b$, is one or even zero.

To keep it simple, we shall assume that all bins are independent,
meaning that
\begin{align}
  p(Q_1,Q_2,\ldots,Q_B\cond\bmy) = \prod_{b=1}^B p(Q_b\cond n_b,\bml_b).
\end{align}
Below, we calculate $p(Q_b\cond n_b,\bml_b)$ for fixed $b$ and hence
drop the subscript, i.e.\ $Q_b\to Q, n_b \to n, N_b\to N$ etc.

%%%%%%%%%%%%%%%%%%%
\fred{Refs: Zech, maybe others, too}

In terms of the generic notation of Section \ref{sec:bayp}, Bayesian
prediction of the distribution of the bin's luminosity
$Q = \sum_{j=1}^n L_j$ given the bin's data $(n,\bml)$ is obtained by
considering not only the likelihood $p(Q\cond\bmtheta_{\rm max})$
using best-fit parameter values $\bmtheta_{\rm max}$, but by
integrating over all possible values of the $\bmtheta$ in terms of the
posterior $p(\bmtheta\cond n,\bml)$,
\begin{align}
  p(Q\cond n,\bml) &= \int d\bmtheta\,p(Q\cond\bmtheta)\,p(\bmtheta\cond n,\bml).
\end{align}
The model's parameters $\bmtheta$ and calculation of this prediction
will be set out in detail below.
